---
title: "01-GO-to-GOslim"
author: "Sam White"
date: "2024-10-30"
output: 
  github_document:
    toc: true
    number_sections: true
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
bibliography: references.bib
---

# INTRO

This notebook is to [take GO IDs and categorize them into their corresponding GOslims](https://github.com/RobertsLab/resources/issues/1989) (GitHub Issue). Specifically, it will use the following tab-delimited input file:

-   [DEGlist_same_2021-2022_forGOslim.tab](https://raw.githubusercontent.com/grace-ac/paper-pycno-sswd-2021-2022/d1cdf13c36085868df4ef4b75d2b7de03ef08d1c/analyses/25-compare-2021-2022/DEGlist_same_2021-2022_forGOslim.tab) (commit: `d1cdf13`)

[Desired output](https://github.com/RobertsLab/resources/issues/1989#issuecomment-2448757444) (GitHub Issue comment) is:

```         
GOslim      Count
GO:0048856  1137
GO:0023052  600
GO:0030154  533
GO:0002376  331
```

This was performed using R, with the following packages:

-   GSEABase [@martinmorgan2017]
-   GO.db [@carlson2017]
-   tidyverse [@wickham2019]

It also relied on gene ontology (GO) information from the [Gene Ontology Consortium](https://geneontology.org/) [@ashburner2000; @thegene2023].

## CODE

```{r setup, include=TRUE}
library(GSEABase)
library(GO.db)
library(knitr)
library(tidyverse)
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

### Variables

```{r set-variables, eval=TRUE}
# Column names corresponding to gene name/ID and GO IDs
GO.ID.column <- "Gene.Ontology.IDs"
gene.ID.column <- "gene_id"

# Relative path or URL to input file
input.file <- "https://raw.githubusercontent.com/grace-ac/paper-pycno-sswd-2021-2022/d1cdf13c36085868df4ef4b75d2b7de03ef08d1c/analyses/25-compare-2021-2022/DEGlist_same_2021-2022_forGOslim.tab"


##### Official GO info - no need to change #####
goslims_obo <- "goslim_generic.obo"
goslims_url <- "http://current.geneontology.org/ontology/subsets/goslim_generic.obo"
```

### Set GSEAbase location and download `goslim_generic.obo`

```{r download-generic-goslim-obo, eval=TRUE}
# Find GSEAbase installation location
gseabase_location <- find.package("GSEABase")

# Load path to GOslim OBO file
goslim_obo_destintation <- file.path(gseabase_location, "extdata", goslims_obo, fsep = "/")

# Download the GOslim OBO file
download.file(url = goslims_url, destfile = goslim_obo_destintation)

# Loads package files
gseabase_files <- system.file("extdata", goslims_obo, package="GSEABase")
```

### Read in gene/GO file

```{r read-in-gene-file, eval=TRUE}
full.gene.df <- read.csv(file = input.file, header = TRUE, sep = "\t")

str(full.gene.df)
```

### Remove rows with NA and remove UniProt IDs

```{r remove-NA-and-uniprotIDs, eval=TRUE}

# Clean whitespace, filter NA/empty rows, select columns, and split GO terms using column name variables
gene.GO.df <- full.gene.df %>%
  mutate(!!GO.ID.column := str_replace_all(.data[[GO.ID.column]], "\\s*;\\s*", ";")) %>% # Clean up spaces around ";"
  filter(!is.na(.data[[gene.ID.column]]) & !is.na(.data[[GO.ID.column]]) & .data[[GO.ID.column]] != "") %>% 
  select(all_of(c(gene.ID.column, GO.ID.column)))


str(gene.GO.df)
```

### "Flatten" gene and GO IDs

```{r flatten-gene-and-GO-IDs, eval=TRUE}
flat.gene.GO.df <- gene.GO.df %>% separate_rows(!!sym(GO.ID.column), sep = ";")

str(flat.gene.GO.df)
```

### Group

```{r group-by-GO, eval=TRUE}
grouped.gene.GO.df <- flat.gene.GO.df %>%
  group_by(!!sym(GO.ID.column)) %>%
  summarise(!!gene.ID.column := paste(.data[[gene.ID.column]], collapse = ","))

str(grouped.gene.GO.df)
```

### Map GO IDs to GOslims

The mapping steps were derived from this [bioconductor forum response](https://support.bioconductor.org/p/128407/#128408)

```{r map-GO-to-GOlims, eval=TRUE}
# Vector of GO IDs
go_ids <- grouped.gene.GO.df[[GO.ID.column]]

str(go_ids)
```

```{r extract-GOslims-from-OBO, eval=TRUE}

# Create GSEAbase GOCollection using `go_ids`
myCollection <- GOCollection(go_ids)

# Retrieve GOslims from GO OBO file set
slim <- getOBOCollection(gseabase_files)

str(slim)
```

```{r retrieve-BP-GOslims, eval=TRUE}
# Retrieve Biological Process (BP) GOslims
slimdf <- goSlim(myCollection, slim, "BP", verbose)
str(slimdf)
```

```{r map-GO-to-GOslims, eval=TRUE}
# List of GOslims and all GO IDs from `go_ids`
gomap <- as.list(GOBPOFFSPRING[rownames(slimdf)])

# Maps `go_ids` to matching GOslims
mapped <- lapply(gomap, intersect, ids(myCollection))

# Append all mapped GO IDs to `slimdf`
# `sapply` needed to apply paste() to create semi-colon delimited values
slimdf$GO.IDs <- sapply(lapply(gomap, intersect, ids(myCollection)), paste, collapse=";")

# Remove "character(0) string from "GO.IDs" column
slimdf$GO.IDs[slimdf$GO.IDs == "character(0)"] <- ""

# Add self-matching GOIDs to "GO.IDs" column, if not present
for (go_id in go_ids) {
  # Check if the go_id is present in the row names
  if (go_id %in% rownames(slimdf)) {
    # Check if the go_id is not present in the GO.IDs column
    # Also removes white space "trimws()" and converts all to upper case to handle
    # any weird, "invisible" formatting issues.
    if (!go_id %in% trimws(toupper(strsplit(slimdf[go_id, "GO.IDs"], ";")[[1]]))) {
      # Append the go_id to the GO.IDs column with a semi-colon separator
      if (length(slimdf$GO.IDs) > 0 && nchar(slimdf$GO.IDs[nrow(slimdf)]) > 0) {
        slimdf[go_id, "GO.IDs"] <- paste0(slimdf[go_id, "GO.IDs"], "; ", go_id)
      } else {
        slimdf[go_id, "GO.IDs"] <- go_id
      }
    }
  }
}

str(slimdf)
```

# Flatten GOslims

```{r flatten-GOslims-file, eval=TRUE}
# "Flatten" file so each row is single GO ID with corresponding GOslim
# rownames_to_column needed to retain row name info
slimdf_separated <- as.data.frame(slimdf %>%
  rownames_to_column('GOslim') %>%
  separate_rows(GO.IDs, sep = ";"))

# Group by unique GO ID
grouped_slimdf <- slimdf_separated %>%
  filter(!is.na(GO.IDs) & GO.IDs != "") %>%
  group_by(GO.IDs) %>%
  summarize(GOslim = paste(GOslim, collapse = ";"),
            Term = paste(Term, collapse = ";"))


str(grouped_slimdf)
```

# Write slimdf to file

## Sort and select slmidf

```{r sort-and-select-slimdf-counts, eval=TRUE}

slimdf.sorted <- slimdf %>% arrange(desc(Count))

slim.count.df <- slimdf.sorted %>% 
  select(Term, Count)

str(slim.count.df)
```

## Write formatted slim.count.df to file

```{r write-GOslim-term-count-to-file, eval=FALSE}

write_tsv(slim.count.df, file = "../output/01-GO-to-GOslim/counts.GOID-per-GOslim_term.tab")
```

# Count unique Biological Process GO IDs

```{r count-unique-BP-GOIDs, eval=TRUE}

# Flatten the list and count total GO IDs
total_go_ids <- length(unlist(gomap))

# Display the total count
total_go_ids

# Unlist to extract all GO IDs, then find unique ones and count them
unique_go_ids <- unique(unlist(gomap))
total_unique_ids <- length(unique_go_ids)

# Display the total count of unique GO IDs
total_unique_ids
```

Total starting BP GO IDs:   `r total_go_ids`

Total unique BP GO IDs:     `r total_unique_ids`

# REFERENCES
