---
title: "Magnetometer Calibration - Unit-A"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
knit: (function(input, ...) {
    out_dir <- file.path(dirname(input), "..", "outputs",
                         tools::file_path_sans_ext(basename(input)));
    dir.create(out_dir, recursive = TRUE, showWarnings = FALSE);
    rmarkdown::render(input, output_dir = out_dir, ...)
  })
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

```{=html}
<!--
Calibration notebook for Unit-A (RP2040 UID df6470a31b8a2f2e).

ALL USER-SETTABLE VALUES live in one place: the `## Variables` chunk under
SETUP below. Unit identity, input paths and the analysis knobs are set
there, as plain R assignments - there is no YAML `params:` block and no
`params = "ask"` prompt, so edit that chunk to point this notebook at
different files.

`log_dir` is the unit's calibration directory; every `calibration-N.csv` in
it is read, where `N` is the magnet count that file was recorded with. The
manifest (`calibration_metadata.csv`) is shared by all three units and is
filtered to `unit_id` here. Every row of every file must carry
UID df6470a31b8a2f2e or the notebook stops.

Paths are relative to this file's directory (`code/`). The `knit:` field
above sends the rendered HTML to `../outputs/calibration_unit-A/` (the
RStudio Knit button honours it; `rmarkdown::render()` does not, so pass
`output_dir = "../outputs/calibration_unit-A"`). CSV exports go to the same
directory via `output_dir` in the Variables chunk.

This notebook is self-contained -- the analysis is inlined below rather than
pulled in from a shared child document, so the file depends on nothing but
its inputs. The trade-off is that the three unit notebooks are now copies:
they are identical apart from the title, this comment and the `## Variables`
chunk, so a change to the analysis has to be made in all three.
`diff calibration_unit-A.Rmd calibration_unit-B.Rmd` should report only
those two blocks.
-->
```

# OVERVIEW

This notebook turns a bench calibration recording into a
**field-strength to distance** curve for one magnetometer unit,
following the approach of the VGWatch valve-gape monitor [@debruyn2024],
whose calibration script fit

$$ d = a \cdot B^{\,b} $$

where $d$ is magnet-to-sensor distance (mm) and $B$ is measured field
magnitude ($\mu$T). That is the model used in the field: the logger
records $B$, and the curve converts it back to a gape distance.

A **hold** is the unit of measurement throughout: the magnet parked at
one distance, with one magnet count, held still while the logger samples
at 1 Hz for at least 60 s (\~61 samples). One row of the calibration
metadata is one hold, and each hold is averaged into a single point on
the curve.

Two models are fit here and compared:

| model | form | notes |
|------------------------|------------------------|------------------------|
| `power` | $d = a B^{b}$ | the published VGWatch form; two parameters |
| `offset` | $d = a B^{b} + c$ | adds an offset for the gap between the magnet's dipole centre / the sensor die and the measured "0 mm" surface. A dipole gives $B \propto r^{-3}$ where $r = d + d_0$, so $c \approx -d_0$. Handles a 0 mm calibration point, which the two-parameter form structurally cannot (it needs $B = \infty$ to predict $d = 0$). |

## Bench protocol this notebook expects

1.  Fix the magnetometer, move the magnet(s) to each distance in turn,
    and **hold each distance still for at least 60 s** while the logger
    runs at 1 Hz - \~61 samples per point.
2.  **Only the holds are recorded.** The logger is not running while the
    magnet is being repositioned, so the recording contains no
    adjustment or settling periods: each hold is a contiguous run of
    samples and the timestamp simply jumps to the next hold. Nothing is
    trimmed off the ends of a hold - every in-window sample is a genuine
    still-magnet reading, and throwing the first and last few away would
    only cost precision.
3.  One file per magnet count, `calibration-N.csv`, where `N` is the
    number of magnets. `calibration-0.csv` is the **no-magnet
    background** recording (magnets well away from the sensor), used to
    subtract the ambient field vector (Earth's field, nearby ferrous
    material) rather than assuming it is zero.
4.  Write down the wall-clock start/end of every hold, then transcribe
    them into the shared calibration metadata CSV (see below). The
    DS3231 RTC timestamps in the log are what the manifest windows are
    matched against.

## Input: the calibration metadata

One row per hold, times in the same `YYYY-MM-DDTHH:MM:SS` form the
logger writes. `calibration/calibration_metadata.csv` holds the rows for
**all** units; this notebook uses only those with `unit == unit_id`.

| column | meaning |
|------------------------------------|------------------------------------|
| `unit` | which logger unit the hold belongs to (`A`, `B`, `C`) |
| `distance` | magnet-to-sensor distance for this hold, mm (ignored on background rows) |
| `magnet.count` | number of magnets on the target (`0` = background hold) |
| `start.time` | wall-clock start of the hold |
| `end.time` | wall-clock end of the hold |
| `data.file` | the `calibration-N.csv` this hold was recorded in |

# SETUP

## Libraries

```{r libraries, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  # Halt on error rather than carrying on with a half-valid document: the
  # UID and manifest checks below are guards, and a guard that only prints
  # its complaint isn't a guard. (`rmarkdown::render` defaults to this
  # anyway; setting it explicitly covers a bare `knitr::knit()` too.)
  error = FALSE,
  fig.width = 10, fig.height = 5
)

library(tidyverse)
library(lubridate)
library(knitr)
```

## Variables

Everything this notebook can be pointed at or tuned with, in one chunk.

```{r variables}
# ---- Unit identity --------------------------------------------------------
# Which logger this notebook is for. `unit_id` selects this unit's rows from
# the shared calibration metadata; `expected_uid` is the RP2040's
# factory-burned UID, checked against every data row so the notebook cannot
# be knit against another unit's recording.
unit_label   <- "Unit-A"
unit_id      <- "A"
expected_uid <- "df6470a31b8a2f2e"

# ---- Inputs ---------------------------------------------------------------
# Paths are relative to this file's directory (`code/`). `log_dir` holds one
# `calibration-N.csv` per magnet count; the manifest covers all three units
# and is filtered to `unit_id` above.
log_dir       <- "../calibration/unit-A"
manifest_path <- "../calibration/calibration_metadata.csv"

# ---- Holds ---------------------------------------------------------------
# No hold trimming. The logger was only running while the magnet was held
# still: each hold is a contiguous run of >= 60 s of samples, and the
# repositioning between holds is simply absent from the recording (the
# timestamps jump). There is no settling period in the data to discard, so
# every in-window sample is used.
#
# Holds with fewer samples than this get flagged - at 1 Hz a 60 s hold
# should give ~61 samples, so anything much below that means the manifest
# times and the recording disagree.
min_samples_per_step <- 45

# ---- Sensor ---------------------------------------------------------------
# TLV493D full-scale range, microtesla (+/- 130 mT). Readings at or above
# `saturation_frac` of this are clipping and are flagged.
sensor_range_uT <- 130000
saturation_frac <- 0.95

# ---- Background -----------------------------------------------------------
# Subtract the ambient field vector measured during `n_magnets = 0` holds.
subtract_background <- TRUE

# ---- Magnet count recommendation ------------------------------------------
# Resolution degrades with distance at a different rate for each magnet
# count, so each count gets its own usable range rather than all being judged
# over one arbitrary window. A count is usable out to the distance where its
# noise-limited resolution first exceeds this value - set it to the smallest
# gape change you need to be able to resolve.
max_acceptable_resolution_mm <- 0.25

# The gape range you actually intend to measure, c(min, max) in mm. The
# headline recommendation is made over this; counts that cannot cover it are
# ruled out rather than ranked.
target_gape_mm <- c(1, 10)

# ---- Misc -----------------------------------------------------------------
deployment_tz <- "America/Los_Angeles"   # timezone the DS3231 was set in
output_dir    <- "../outputs/calibration_unit-A"
```

```{r output-dir}
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

cat(sprintf(
  "Unit:          %s (%s)\nExpected UID:  %s\nLog dir:       %s\nManifest:      %s\nOutput dir:    %s\n",
  unit_label, unit_id, expected_uid,
  log_dir, manifest_path, normalizePath(output_dir)
))
```

# DATA IMPORT

## Read the logger CSV(s)

One file per magnet count - `calibration-0.csv` (background) through
`calibration-3.csv` - all read together. `source_file` carries the file
name through, because that is what the metadata's `data.file` column
matches against: two units recorded on the same day can share wall-clock
times, so time alone is not enough to pin a sample to a hold.

```{r read-log}
stopifnot("`log_dir` does not exist." = dir.exists(log_dir))

log_files <- list.files(log_dir, pattern = "^calibration-[0-9]+\\.csv$",
                        full.names = TRUE)

stopifnot("No `calibration-N.csv` files found in `log_dir`." = length(log_files) > 0)

raw <- log_files %>%
  set_names(basename(.)) %>%
  map(
    read_csv,
    col_types = cols(
      timestamp  = col_character(),
      rp2040_uid = col_character(),
      x_uT       = col_double(),
      y_uT       = col_double(),
      z_uT       = col_double()
    )
  ) %>%
  list_rbind(names_to = "source_file") %>%
  mutate(sample_id = row_number(), .before = everything())

glimpse(raw)
```

## Confirm the data came from this unit

The RP2040's factory-burned UID is written on every data row, so a log
can always be traced back to the board that recorded it. Knitting stops
here if the file doesn't belong to this unit - that is the whole point
of having one notebook per unit.

Note the UID identifies the **Adalogger board only**. The TLV493D
exposes no unique ID over I2C, so if magnetometers get swapped between
boards the board-to-sensor pairing has to be tracked by physically
marking the parts.

```{r verify-uid}
uids <- sort(unique(raw$rp2040_uid))

if (length(uids) != 1) {
  stop(sprintf(
    "Expected exactly one rp2040_uid in the input, found %d: %s",
    length(uids), paste(uids, collapse = ", ")
  ))
}

if (!identical(uids, expected_uid)) {
  stop(sprintf(
    paste0("UID MISMATCH - this data was NOT recorded by %s.\n",
           "  expected: %s\n  found:    %s\n",
           "Knit the notebook for the unit with the matching UID instead."),
    unit_label, expected_uid, uids
  ))
}

cat(sprintf("UID confirmed: %s recorded by %s (%d samples).\n",
            uids, unit_label, nrow(raw)))
```

## Parse timestamps

Matching samples to manifest windows needs real wall-clock time, so a
session that fell back to `T+123.456s` elapsed markers (RTC not detected
at boot - look for `RTC_INIT_ERROR` in `eventlog.csv`) can't be
calibrated against a hand-written time log. That is a hard stop.

```{r parse-timestamps}
log_data <- raw %>%
  mutate(timestamp = ymd_hms(timestamp, tz = deployment_tz, quiet = TRUE))

n_unparsed <- sum(is.na(log_data$timestamp))

if (n_unparsed > 0) {
  stop(sprintf(
    paste0("%d of %d rows have no absolute RTC timestamp (elapsed-time ",
           "fallback). Calibration needs wall-clock time to match the ",
           "manifest windows - re-record with a working DS3231."),
    n_unparsed, nrow(log_data)
  ))
}

cat(sprintf("Recording spans %s to %s (%.1f min).\n",
            format(min(log_data$timestamp)), format(max(log_data$timestamp)),
            as.numeric(difftime(max(log_data$timestamp),
                                min(log_data$timestamp), units = "mins"))))
```

## Raw field magnitude

Magnitude `sqrt(x^2 + y^2 + z^2)` is rotation-invariant, so it doesn't
depend on exactly how the sensor is oriented relative to the magnet -
the same reason it is the gape proxy in `gaping_analysis.Rmd`.

```{r raw-magnitude}
log_data <- log_data %>%
  mutate(field_raw_uT = sqrt(x_uT^2 + y_uT^2 + z_uT^2))

log_data %>%
  select(x_uT, y_uT, z_uT, field_raw_uT) %>%
  summary()
```

## Read the calibration manifest

The metadata file covers every unit, so it is filtered to this one
first. Column names are mapped to the names used throughout the rest of
the notebook (`magnet.count` -\> `n_magnets`, `distance` -\>
`distance_mm`, and so on), and `distance_mm` is blanked on background
rows where it has no meaning.

```{r read-manifest}
stopifnot("`manifest_path` does not exist." = file.exists(manifest_path))

manifest_all <- read_csv(
  manifest_path,
  col_types = cols(
    unit         = col_character(),
    distance     = col_double(),
    magnet.count = col_integer(),
    start.time   = col_character(),
    end.time     = col_character(),
    data.file    = col_character()
  )
)

manifest <- manifest_all %>%
  filter(unit == unit_id) %>%
  transmute(
    n_magnets   = magnet.count,
    distance_mm = if_else(magnet.count > 0, distance, NA_real_),
    start_time  = ymd_hms(start.time, tz = deployment_tz, quiet = TRUE),
    end_time    = ymd_hms(end.time,   tz = deployment_tz, quiet = TRUE),
    source_file = data.file
  ) %>%
  arrange(start_time) %>%
  mutate(step_id = row_number(), .before = everything())

if (nrow(manifest) == 0) {
  stop(sprintf(
    "No rows with unit == '%s' in %s - units present: %s",
    unit_id, manifest_path,
    paste(sort(unique(manifest_all$unit)), collapse = ", ")
  ))
}

manifest %>%
  mutate(hold_s = as.numeric(difftime(end_time, start_time, units = "secs"))) %>%
  kable(caption = sprintf("Calibration metadata for %s", unit_label))
```

## Validate the manifest

Catches the transcription mistakes that are easy to make at the bench
and hard to spot later: unparsed times, backwards windows, overlapping
holds, missing distances on magnet rows, and a `data.file` that names a
file which was not read.

```{r validate-manifest}
bad_times <- manifest %>% filter(is.na(start_time) | is.na(end_time))
backwards <- manifest %>% filter(end_time <= start_time)
no_dist   <- manifest %>% filter(n_magnets > 0, is.na(distance_mm))
no_file   <- manifest %>% filter(!source_file %in% basename(log_files))

overlaps <- manifest %>%
  arrange(source_file, start_time) %>%
  filter(source_file == lag(source_file), start_time < lag(end_time)) %>%
  select(step_id, n_magnets, distance_mm, start_time, end_time, source_file)

problems <- c(
  if (nrow(bad_times)) sprintf("%d row(s) with unparseable start/end time", nrow(bad_times)),
  if (nrow(backwards)) sprintf("%d row(s) where end_time <= start_time", nrow(backwards)),
  if (nrow(no_dist))   sprintf("%d magnet row(s) missing distance_mm", nrow(no_dist)),
  if (nrow(no_file))   sprintf("%d row(s) naming a data.file not present in log_dir: %s",
                               nrow(no_file), paste(unique(no_file$source_file), collapse = ", ")),
  if (nrow(overlaps))  sprintf("%d overlapping time window(s)", nrow(overlaps))
)

if (length(problems) > 0) {
  stop("Manifest problems:\n  - ", paste(problems, collapse = "\n  - "))
}

cat(sprintf("Manifest OK: %d hold(s) - %d background, %d magnet holds across %s magnet count(s).\n",
            nrow(manifest),
            sum(manifest$n_magnets == 0),
            sum(manifest$n_magnets > 0),
            paste(sort(unique(manifest$n_magnets[manifest$n_magnets > 0])), collapse = "/")))
```

## Assign samples to holds

Samples are matched to holds **within a file**: a sample can only belong
to a hold that the metadata says was recorded in the file that sample
came from.

```{r assign-samples}
assigned <- log_data %>%
  inner_join(manifest %>% select(step_id, n_magnets, distance_mm,
                                 start_time, end_time, source_file),
             by = "source_file", relationship = "many-to-many") %>%
  filter(timestamp >= start_time, timestamp <= end_time)

# A sample landing in two windows means the manifest is internally
# inconsistent in a way the overlap check above didn't catch.
multi_assigned <- assigned %>% count(sample_id) %>% filter(n > 1)
stopifnot("Some samples fall inside more than one manifest window." =
            nrow(multi_assigned) == 0)

cat(sprintf(
  "%d of %d samples (%.1f%%) fall inside a manifest window; %d are unclaimed.\n",
  nrow(assigned), nrow(log_data), 100 * nrow(assigned) / nrow(log_data),
  nrow(log_data) - nrow(assigned)
))
```

## QC: do the manifest windows line up with the trace?

The single most useful plot here. Grey is the whole recording - a
staircase, one tread per hold, with no risers because the logger was off
while the magnet was moved. Coloured points are the samples the metadata
actually claimed. Every tread should be fully coloured and flat: grey
showing through, or a coloured run that steps part-way through, means
those hand-written times are off and need fixing before anything below
is worth reading.

```{r plot-coverage, fig.height = 5}
ggplot(log_data, aes(x = timestamp, y = field_raw_uT)) +
  geom_line(colour = "grey75", linewidth = 0.3) +
  geom_point(data = assigned, aes(colour = factor(n_magnets)), size = 0.5) +
  scale_y_log10() +
  scale_colour_viridis_d(name = "Magnets", end = 0.9) +
  labs(
    title = paste(unit_label, "- calibration recording vs. manifest windows"),
    subtitle = "Grey = full recording; coloured = samples claimed by a manifest hold",
    x = "Time", y = expression(paste("Raw field magnitude (", mu, "T, log scale)"))
  ) +
  theme_bw()
```

To inspect one stretch closely, add a `coord_cartesian(xlim = ...)` to
the plot above rather than re-filtering the data.

## Hold completeness

**No samples are trimmed.** Earlier versions of this notebook discarded
a settling window at each end of every hold, because the recording then
ran continuously and the ends of a window caught the hand still coming
to rest or already moving away. These recordings do not work that way:
logging covered the holds only, each at least 60 s long, with the
repositioning simply absent from the file. There is no settling period
in the data to remove, so every in-window sample is kept and the
per-hold means below are built from \~61 samples rather than \~45.

What is still worth checking is that each hold got the samples it should
have, and that no sample went unclaimed - both are symptoms of
hand-written times that drifted from the recording.

```{r hold-coverage}
hold_samples <- assigned

hold_counts <- hold_samples %>%
  count(step_id, n_magnets, distance_mm, name = "n_samples") %>%
  left_join(
    manifest %>%
      mutate(hold_s = as.numeric(difftime(end_time, start_time, units = "secs"))) %>%
      select(step_id, hold_s),
    by = "step_id"
  )

thin_steps <- hold_counts %>% filter(n_samples < min_samples_per_step)

if (nrow(thin_steps) > 0) {
  thin_steps %>%
    kable(caption = sprintf("WARNING - holds with fewer than %d samples",
                            min_samples_per_step)) %>%
    print()
}

unclaimed <- log_data %>% filter(!sample_id %in% hold_samples$sample_id)

if (nrow(unclaimed) > 0) {
  unclaimed %>%
    count(source_file, name = "n_unclaimed") %>%
    kable(caption = paste("Samples in the recording that no metadata window claims",
                          "- check these times")) %>%
    print()
}

cat(sprintf(
  "Using all %d in-window samples across %d hold(s): %.0f-%.0f samples per hold (median %.0f).\n",
  nrow(hold_samples), nrow(hold_counts),
  min(hold_counts$n_samples), max(hold_counts$n_samples),
  median(hold_counts$n_samples)
))
```

## Saturation check

The TLV493D clips at its full-scale range. A clipped reading looks like
a flat-topped plateau and will quietly bend the low-distance end of the
curve, so any hold with near-full-scale readings is flagged here and
should be dropped or re-recorded with fewer/weaker magnets.

```{r saturation-check}
sat_limit <- sensor_range_uT * saturation_frac

saturated <- hold_samples %>%
  mutate(any_axis_max = pmax(abs(x_uT), abs(y_uT), abs(z_uT))) %>%
  group_by(n_magnets, distance_mm) %>%
  summarise(
    max_axis_uT = max(any_axis_max),
    n_saturated = sum(any_axis_max >= sat_limit),
    .groups = "drop"
  ) %>%
  filter(n_saturated > 0)

if (nrow(saturated) > 0) {
  saturated %>%
    kable(caption = sprintf(
      "SATURATED holds (>= %.0f uT on an axis) - exclude these before trusting the fit",
      sat_limit))
} else {
  cat(sprintf("No saturation: all axes stayed below %.0f uT (%.0f%% of the +/-%.0f uT range).\n",
              sat_limit, 100 * saturation_frac, sensor_range_uT))
}
```

# BACKGROUND CORRECTION

Ambient field (Earth's \~50 $\mu$T, plus anything ferrous on the bench)
adds as a **vector**, not as a magnitude, so it is subtracted per axis
before magnitude is recomputed. This matters most at the far end of the
sweep, where the magnet's own contribution is smallest.

```{r background-subtraction}
bg_samples <- hold_samples %>% filter(n_magnets == 0)

background <- if (subtract_background && nrow(bg_samples) > 0) {
  bg_samples %>% summarise(bx = mean(x_uT), by = mean(y_uT), bz = mean(z_uT))
} else {
  tibble(bx = 0, by = 0, bz = 0)
}

if (!subtract_background) {
  cat("Background subtraction disabled (`subtract_background = FALSE`).\n")
} else if (nrow(bg_samples) == 0) {
  cat("No `n_magnets = 0` holds in the manifest - no background subtracted.\n")
} else {
  cat(sprintf(
    "Background from %d sample(s): x = %.2f, y = %.2f, z = %.2f uT (|B| = %.2f uT).\n",
    nrow(bg_samples), background$bx, background$by, background$bz,
    sqrt(background$bx^2 + background$by^2 + background$bz^2)
  ))
}

calib <- hold_samples %>%
  filter(n_magnets > 0) %>%
  mutate(
    x_corr   = x_uT - background$bx,
    y_corr   = y_uT - background$by,
    z_corr   = z_uT - background$bz,
    field_uT = sqrt(x_corr^2 + y_corr^2 + z_corr^2),
    n_magnets = factor(n_magnets)
  )
```

# PER-HOLD SUMMARIES

## Summary table

`cv_pct` is the within-hold coefficient of variation - a stability check
on the hold itself. A few percent is normal; a large value means the
magnet moved, or the readings were clipping.

```{r step-summary}
step_summary <- calib %>%
  group_by(n_magnets, distance_mm) %>%
  summarise(
    n           = n(),
    mean_uT     = mean(field_uT),
    sd_uT       = sd(field_uT),
    cv_pct      = 100 * sd_uT / mean_uT,
    min_uT      = min(field_uT),
    max_uT      = max(field_uT),
    .groups     = "drop"
  ) %>%
  arrange(n_magnets, distance_mm)

step_summary %>%
  mutate(across(c(mean_uT, sd_uT, min_uT, max_uT), ~ round(.x, 1)),
         cv_pct = round(cv_pct, 2)) %>%
  kable(caption = "Field magnitude per calibration hold (background-corrected)")
```

## Within-hold spread

```{r plot-step-stability, fig.height = 4}
ggplot(calib, aes(x = factor(distance_mm), y = field_uT, fill = n_magnets)) +
  geom_boxplot(outlier.size = 0.4, linewidth = 0.3) +
  scale_y_log10() +
  scale_fill_viridis_d(name = "Magnets") +
  facet_grid(cols = vars(n_magnets), labeller = label_both) +
  labs(
    title = paste(unit_label, "- spread of readings within each hold"),
    x = "Distance to magnet (mm)",
    y = expression(paste("Field magnitude (", mu, "T, log scale)"))
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

# MODEL FITTING

Fits use the per-hold **means**, one point per distance, matching the
published VGWatch calibration. Sample-level points are still shown on
the plots below, but fitting on means keeps a noisy hold from dominating
the fit just because it has spread.

## Fit both models per magnet count

Starting values matter for `nls`. `power` starts from a log-log linear
regression (exact if the relationship were a pure power law); `offset`
starts from the dipole expectation $b = -1/3$, with $a$ and $c$ taken
from a linear regression of $d$ on $B^{-1/3}$.

```{r fit-models}
fit_power <- function(df) {
  pos <- df %>% filter(distance_mm > 0)
  start <- coef(lm(log(distance_mm) ~ log(mean_uT), data = pos))
  nls(
    distance_mm ~ a * mean_uT^b, data = df,
    start   = list(a = unname(exp(start[1])), b = unname(start[2])),
    control = nls.control(maxiter = 200, warnOnly = TRUE)
  )
}

fit_offset <- function(df) {
  lin <- lm(distance_mm ~ I(mean_uT^(-1 / 3)), data = df)
  nls(
    distance_mm ~ a * mean_uT^b + c, data = df,
    start   = list(a = unname(coef(lin)[2]), b = -1 / 3, c = unname(coef(lin)[1])),
    control = nls.control(maxiter = 200, warnOnly = TRUE)
  )
}

safely_fit <- function(f) function(df) tryCatch(f(df), error = function(e) NULL)

fits <- step_summary %>%
  nest(data = -n_magnets) %>%
  mutate(
    power  = map(data, safely_fit(fit_power)),
    offset = map(data, safely_fit(fit_offset))
  ) %>%
  pivot_longer(c(power, offset), names_to = "model", values_to = "fit")

failed <- fits %>% filter(map_lgl(fit, is.null))
if (nrow(failed) > 0) {
  cat("Models that failed to converge:\n")
  failed %>% select(n_magnets, model) %>% kable() %>% print()
}

fits <- fits %>% filter(!map_lgl(fit, is.null))
stopifnot("No models converged - check the data above." = nrow(fits) > 0)
```

## Coefficients

```{r coefficient-table}
coefficients <- fits %>%
  mutate(tidied = map(fit, ~ as_tibble(t(coef(.x))))) %>%
  select(n_magnets, model, tidied) %>%
  unnest(tidied) %>%
  arrange(n_magnets, model)

# `power` has no `c` term; keep the column present so downstream code that
# reports/uses it doesn't have to branch on which models converged.
if (!"c" %in% names(coefficients)) coefficients$c <- NA_real_

coefficients %>%
  mutate(across(where(is.numeric), ~ signif(.x, 5))) %>%
  kable(caption = "Fitted coefficients (distance_mm = a * field_uT^b [+ c])")
```

## Model comparison

`rmse_mm` is the root-mean-square distance error across the calibration
points - read it as "how many mm off this curve is, on average, over the
range it was fit on".

```{r model-comparison}
model_stats <- fits %>%
  mutate(
    n_points = map_int(fit, ~ length(residuals(.x))),
    rmse_mm  = map_dbl(fit, ~ sqrt(mean(residuals(.x)^2))),
    max_err_mm = map_dbl(fit, ~ max(abs(residuals(.x)))),
    aic      = map_dbl(fit, AIC)
  ) %>%
  select(n_magnets, model, n_points, rmse_mm, max_err_mm, aic) %>%
  group_by(n_magnets) %>%
  mutate(delta_aic = aic - min(aic)) %>%
  ungroup() %>%
  arrange(n_magnets, aic)

model_stats %>%
  mutate(across(c(rmse_mm, max_err_mm, aic, delta_aic), ~ round(.x, 3))) %>%
  kable(caption = "Model comparison per magnet count (lower AIC / RMSE is better)")

best_models <- model_stats %>%
  group_by(n_magnets) %>%
  slice_min(aic, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(n_magnets, best_model = model, rmse_mm)

best_models %>% kable(caption = "Best-fitting model per magnet count")
```

## Magnet count is part of the calibration

A curve maps field strength to distance **for one particular magnet
configuration**. Adding magnets raises the field at every distance, so
the same reading means a different distance - applying the wrong curve
produces a confidently wrong number, not an obviously broken one.

For a point dipole, $B = n k /(d + d_0)^3$, so the inverted form
$d = a B^{b} + c$ keeps the same $b$ and $c$ while $a$ scales as
$n^{1/3}$. Checking that against the fitted values tests whether the
magnets really are stacking additively. Deviation is informative rather
than alarming: a stack has physical length, so its effective dipole
centre sits further from the leading face than a single magnet's does,
which shifts both $a$ and $c$. A *large* deviation, though, points at a
mis-seated or reversed magnet in the stack, and is worth chasing before
trusting the fit.

```{r magnet-scaling}
scaling <- coefficients %>%
  inner_join(best_models %>% select(n_magnets, best_model), by = "n_magnets") %>%
  filter(model == best_model) %>%
  mutate(n = as.numeric(as.character(n_magnets))) %>%
  arrange(n)

if (!1 %in% scaling$n) {
  cat("No single-magnet holds in this session - skipping the additivity check.\n")
} else {
  scaling %>%
    mutate(
      a_ratio_observed = a / a[n == 1],
      a_ratio_expected = n^(1 / 3),
      deviation_pct    = 100 * (a_ratio_observed / a_ratio_expected - 1)
    ) %>%
    select(n_magnets, model, a, a_ratio_observed, a_ratio_expected,
           deviation_pct, b, c) %>%
    mutate(across(where(is.numeric), ~ signif(.x, 4))) %>%
    kable(caption = "Additivity check: does `a` scale as n^(1/3), with b and c shared?")
}
```

```{r wrong-curve-cost}
# Concrete cost of picking the wrong row out of the exported coefficients:
# take a few true distances, work out what field each magnet count actually
# produces there, then read that field back through every curve. Probe
# distances are taken from the calibrated range rather than hard-coded, so
# nothing here is an extrapolation past the sweep that was actually run.
probe_mm <- unique(round(seq(min(step_summary$distance_mm),
                             max(step_summary$distance_mm), length.out = 4)))

wrong_curve <- scaling %>%
  select(n_true = n, a, b, c) %>%
  cross_join(tibble(true_distance_mm = probe_mm)) %>%
  mutate(field_uT = ((true_distance_mm - coalesce(c, 0)) / a)^(1 / b)) %>%
  select(n_true, true_distance_mm, field_uT) %>%
  cross_join(scaling %>% select(n_curve = n, a, b, c)) %>%
  mutate(read_distance_mm = round(a * field_uT^b + coalesce(c, 0), 2)) %>%
  select(n_true, true_distance_mm, n_curve, read_distance_mm) %>%
  pivot_wider(names_from = n_curve, values_from = read_distance_mm,
              names_prefix = "read_with_n=") %>%
  arrange(n_true, true_distance_mm)

wrong_curve %>%
  kable(caption = paste("Distance read when the applied curve's magnet count",
                        "does not match the data's (diagonal = correct)"))
```

Read the off-diagonal entries as the error from a metadata mix-up. Note
that a mismatched curve can even return a negative distance - the one
failure mode here that announces itself.

## Calibration curves

Axes follow the published figure: measured field on x, inferred distance
on y - the direction the curve is actually used in.

```{r plot-curves, fig.height = 4.5}
curve_grid <- fits %>%
  mutate(
    pred = map2(fit, data, function(f, d) {
      grid <- tibble(mean_uT = exp(seq(log(min(d$mean_uT)), log(max(d$mean_uT)),
                                       length.out = 200)))
      grid %>% mutate(distance_mm = predict(f, newdata = grid))
    })
  ) %>%
  select(n_magnets, model, pred) %>%
  unnest(pred)

annot <- coefficients %>%
  mutate(label = sprintf("%s: a=%.3g, b=%.4f%s", model, a, b,
                         if_else(is.na(c), "", sprintf(", c=%.3g", c)))) %>%
  group_by(n_magnets) %>%
  summarise(label = paste(label, collapse = "\n"), .groups = "drop")

ggplot(step_summary, aes(x = mean_uT, y = distance_mm)) +
  geom_point(size = 2) +
  geom_errorbar(aes(xmin = mean_uT - sd_uT, xmax = mean_uT + sd_uT),
                orientation = "y", width = 0.2) +
  geom_line(data = curve_grid, aes(colour = model), linewidth = 0.7) +
  geom_text(data = annot, aes(x = Inf, y = Inf, label = label),
            hjust = 1.05, vjust = 1.3, size = 2.6, inherit.aes = FALSE) +
  scale_x_log10() +
  scale_colour_brewer(palette = "Set1", name = "Model") +
  facet_grid(cols = vars(n_magnets), labeller = label_both) +
  labs(
    title = paste(unit_label, "- calibration curves"),
    subtitle = "Points are per-hold means +/- 1 SD",
    x = expression(paste("Magnetic field strength (", mu, "T, log scale)")),
    y = "Distance to magnet (mm)"
  ) +
  theme_bw()
```

## Residuals

Systematic curvature in these residuals (rather than scatter about zero)
is the sign that the model form is wrong over part of the range -
typically the two-parameter `power` form struggling near 0 mm.

```{r plot-residuals, fig.height = 4}
residual_data <- fits %>%
  mutate(res = map2(fit, data, ~ .y %>% mutate(residual_mm = residuals(.x)))) %>%
  select(n_magnets, model, res) %>%
  unnest(res)

ggplot(residual_data, aes(x = distance_mm, y = residual_mm, colour = model)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(size = 2) +
  geom_line(linewidth = 0.4) +
  scale_colour_brewer(palette = "Set1", name = "Model") +
  facet_grid(cols = vars(n_magnets), labeller = label_both) +
  labs(
    title = paste(unit_label, "- fit residuals"),
    x = "Distance to magnet (mm)",
    y = "Observed - predicted (mm)"
  ) +
  theme_bw()
```

# CHOOSING A MAGNET COUNT

More magnets give a stronger signal but saturate sooner at close range.
The useful quantity is **resolution**: how many $\mu$T of change there
is per mm of gape, relative to the sensor's own noise. `mm_per_uT_noise`
below is the smallest distance change distinguishable from within-hold
noise at each point - smaller is better.

Resolution degrades with distance, and it degrades at a *different rate*
for each magnet count - so there is no single sweep range that is fair
to all three. Judging every count over one window either flatters the
weak ones (if the window is short) or penalises them for holds nobody
would act on (if it is long). The assessment therefore runs in three
steps.

**1. How far is each count usable?** A count is usable out to the
distance where its resolution first crosses
`max_acceptable_resolution_mm` (currently
`r max_acceptable_resolution_mm` mm). One magnet might be good to 11 mm
and three to 14 mm; that difference is a result in itself, not noise to
be averaged away.

**2. Re-assess each count over its own range.** Scoring a count on holds
past its usable limit judges it on data you would never act on, so each
count is summarised over its own range only.

**3. Which count wins over the range you care about?** Counts that
cannot cover `target_gape_mm` are ruled out rather than ranked, and the
survivors compared over exactly that range. A per-band table shows how
the answer shifts with the range, because a count that wins over 1-5 mm
need not win over 1-20 mm.

A median can also hide a tie, so the margin to the runner-up is printed,
and if ranking on worst-case resolution would pick a different count
that is said out loud rather than quietly discarded.

```{r magnet-resolution}
resolution <- step_summary %>%
  arrange(n_magnets, distance_mm) %>%
  group_by(n_magnets) %>%
  mutate(
    d_uT_per_mm     = abs(mean_uT - lag(mean_uT)) / (distance_mm - lag(distance_mm)),
    mm_per_uT_noise = sd_uT / d_uT_per_mm
  ) %>%
  ungroup()

resolution %>%
  filter(!is.na(d_uT_per_mm)) %>%
  select(n_magnets, distance_mm, mean_uT, sd_uT, d_uT_per_mm, mm_per_uT_noise) %>%
  mutate(across(where(is.numeric), ~ signif(.x, 4))) %>%
  kable(caption = "Sensitivity and noise-limited resolution between adjacent holds")
```

```{r plot-resolution, fig.height = 4}
resolution %>%
  filter(!is.na(mm_per_uT_noise)) %>%
  ggplot(aes(x = distance_mm, y = mm_per_uT_noise, colour = n_magnets)) +
  geom_hline(yintercept = max_acceptable_resolution_mm,
             linetype = "dashed", colour = "firebrick") +
  geom_line(linewidth = 0.6) +
  geom_point(size = 2) +
  scale_y_log10() +
  scale_colour_viridis_d(name = "Magnets") +
  labs(
    title = paste(unit_label, "- noise-limited distance resolution"),
    subtitle = sprintf(paste("Lower is better. Dashed line = the %g mm acceptance",
                             "limit;\nwhere each curve crosses it is that count's",
                             "usable limit."),
                       max_acceptable_resolution_mm),
    x = "Distance to magnet (mm)", y = "Resolution (mm, log scale)"
  ) +
  theme_bw()
```

## Step 1 - how far is each magnet count usable?

```{r usable-range}
# Peak field comes from `step_summary`, not `resolution`: `resolution` has no
# row for the closest hold (there is no preceding hold to difference against),
# and that closest hold is exactly where the field peaks and saturation risk
# is highest.
field_range <- step_summary %>%
  group_by(n_magnets) %>%
  summarise(max_field_uT = max(mean_uT), .groups = "drop") %>%
  mutate(pct_of_range = 100 * max_field_uT / sensor_range_uT)

res_ok       <- resolution %>% filter(!is.na(mm_per_uT_noise))
sweep_min_mm <- min(step_summary$distance_mm)

# `mm_per_uT_noise` at distance d describes the step from the previous hold
# up to d, so a count is usable out to the last d BEFORE the first crossing.
usable_to <- function(d, res, limit) {
  bad <- which(res > limit)
  if (length(bad) == 0) return(max(d))   # never crosses - good to the end
  if (bad[1] == 1)      return(NA_real_) # too coarse from the very first step
  d[bad[1] - 1]
}

usable_range <- res_ok %>%
  arrange(n_magnets, distance_mm) %>%
  group_by(n_magnets) %>%
  summarise(
    usable_max_mm = usable_to(distance_mm, mm_per_uT_noise,
                              max_acceptable_resolution_mm),
    swept_to_mm   = max(distance_mm),
    .groups = "drop"
  ) %>%
  mutate(usable_range = if_else(
    is.na(usable_max_mm), "none",
    sprintf("%g-%g mm", sweep_min_mm, usable_max_mm)))

usable_range %>%
  left_join(field_range, by = "n_magnets") %>%
  select(n_magnets, usable_range, usable_max_mm, swept_to_mm,
         max_field_uT, pct_of_range) %>%
  mutate(across(where(is.numeric), ~ signif(.x, 4))) %>%
  kable(caption = sprintf(
    "Usable range per magnet count (resolution stays under %g mm)",
    max_acceptable_resolution_mm))

cat(sprintf("Swept to %g mm. Usable limits: %s\n",
            max(res_ok$distance_mm),
            paste(sprintf("%s magnet(s) to %s",
                          as.character(usable_range$n_magnets),
                          if_else(is.na(usable_range$usable_max_mm), "(none)",
                                  sprintf("%g mm", usable_range$usable_max_mm))),
                  collapse = "; ")))
```

## Step 2 - each count scored over its own usable range

```{r per-count-assessment}
per_count <- res_ok %>%
  inner_join(usable_range %>% select(n_magnets, usable_max_mm, usable_range),
             by = "n_magnets") %>%
  filter(!is.na(usable_max_mm), distance_mm <= usable_max_mm) %>%
  group_by(n_magnets, usable_range) %>%
  summarise(
    median_resolution_mm = median(mm_per_uT_noise),
    worst_resolution_mm  = max(mm_per_uT_noise),
    n_steps              = n(),
    .groups = "drop"
  ) %>%
  arrange(median_resolution_mm)

if (nrow(per_count) == 0) {
  cat(sprintf(paste0("No magnet count keeps resolution under %g mm even at the ",
                     "closest holds. Loosen `max_acceptable_resolution_mm`, or ",
                     "re-record with stronger magnets.\n"),
              max_acceptable_resolution_mm))
} else {
  per_count %>%
    mutate(across(where(is.numeric), ~ signif(.x, 4))) %>%
    kable(caption = paste("Each count over ITS OWN usable range - the ranges",
                          "differ, so these are not directly comparable"))
}
```

Each row above is scored over a **different** range, so a lower median
here does not by itself mean a better count - a count that gives up at
11 mm is being flattered relative to one still going at 14 mm. Step 3
makes the comparison on equal terms.

## Step 3 - which count is best over a given range?

```{r magnet-recommendation}
# Resolution at distance d covers the step (d-1, d], so a range [lo, hi]
# is the steps with d > lo and d <= hi.
rank_over <- function(lo, hi) {
  res_ok %>%
    filter(distance_mm > lo, distance_mm <= hi) %>%
    group_by(n_magnets) %>%
    summarise(median_resolution_mm = median(mm_per_uT_noise),
              worst_resolution_mm  = max(mm_per_uT_noise),
              .groups = "drop") %>%
    left_join(usable_range %>% select(n_magnets, usable_max_mm),
              by = "n_magnets") %>%
    mutate(covers = !is.na(usable_max_mm) & usable_max_mm >= hi) %>%
    arrange(median_resolution_mm)
}

band_max <- sort(unique(c(5, 10, 15, max(res_ok$distance_mm))))
band_max <- band_max[band_max <= max(res_ok$distance_mm)]

bands <- map_dfr(band_max, function(h) {
  r <- rank_over(sweep_min_mm, h) %>% filter(covers)
  tibble(
    range          = sprintf("%g-%g mm", sweep_min_mm, h),
    usable_counts  = nrow(r),
    best_n_magnets = if (nrow(r) == 0) NA_character_ else as.character(r$n_magnets[1]),
    median_res_mm  = if (nrow(r) == 0) NA_real_ else r$median_resolution_mm[1],
    runner_up      = if (nrow(r) > 1)
      sprintf("%s magnet(s) at %.3f mm", as.character(r$n_magnets[2]),
              r$median_resolution_mm[2]) else "-"
  )
})

bands %>%
  mutate(across(where(is.numeric), ~ signif(.x, 4))) %>%
  kable(caption = paste("Best magnet count per range, among counts that stay",
                        "under the resolution limit that far"))

tgt_lo <- target_gape_mm[1]
tgt_hi <- target_gape_mm[2]
eligible <- rank_over(tgt_lo, tgt_hi) %>% filter(covers)

if (nrow(eligible) == 0) {
  reach <- usable_range %>% filter(!is.na(usable_max_mm))
  # Nothing qualifies, but downstream chunks still need a count to
  # demonstrate with: fall back to whichever reaches furthest.
  recommended_n_magnets <- {
    r <- usable_range %>% filter(!is.na(usable_max_mm))
    if (nrow(r) == 0) as.character(sort(unique(res_ok$n_magnets))[1])
    else as.character(r$n_magnets[which.max(r$usable_max_mm)])
  }

  cat(sprintf(
    paste0("NO magnet count holds %g mm resolution all the way to %g mm ",
           "(`target_gape_mm`).\n  Furthest any count reaches: %s.\n",
           "  Either shorten `target_gape_mm`, loosen ",
           "`max_acceptable_resolution_mm`, or re-record with more magnets.\n"),
    max_acceptable_resolution_mm, tgt_hi,
    if (nrow(reach) == 0) "none" else
      sprintf("%g mm with %s magnet(s)", max(reach$usable_max_mm),
              as.character(reach$n_magnets[which.max(reach$usable_max_mm)]))))
} else {
  best_n  <- as.character(eligible$n_magnets[1])
  recommended_n_magnets <- best_n
  best_fr <- field_range %>% filter(as.character(n_magnets) == best_n)

  cat(sprintf(
    "Suggested for %s over %g-%g mm: %s magnet(s) - median resolution %.3f mm, worst %.3f mm, peak field %.0f uT (%.1f%% of range).\n",
    unit_label, tgt_lo, tgt_hi, best_n,
    eligible$median_resolution_mm[1], eligible$worst_resolution_mm[1],
    best_fr$max_field_uT[1], best_fr$pct_of_range[1]))

  ruled_out <- rank_over(tgt_lo, tgt_hi) %>% filter(!covers)
  if (nrow(ruled_out) > 0) {
    cat(sprintf("  Ruled out (usable range too short to cover %g mm): %s.\n",
                tgt_hi,
                paste(sprintf("%s magnet(s), good to %s",
                              as.character(ruled_out$n_magnets),
                              if_else(is.na(ruled_out$usable_max_mm), "(none)",
                                      sprintf("%g mm", ruled_out$usable_max_mm))),
                      collapse = "; ")))
  }

  # A median can separate two counts by a fraction of a percent, which is a
  # tie, not a result.
  if (nrow(eligible) > 1) {
    margin_pct <- 100 * (eligible$median_resolution_mm[2] /
                         eligible$median_resolution_mm[1] - 1)
    runner_n <- as.character(eligible$n_magnets[2])
    cat(sprintf("  Runner-up: %s magnet(s) at %.3f mm - %.1f%% behind.\n",
                runner_n, eligible$median_resolution_mm[2], margin_pct))
    if (margin_pct < 10) {
      cat(sprintf(
        paste0("  That is effectively a tie over this range: %s and %s ",
               "magnet(s) resolve gape equally well here, so choose on magnet ",
               "weight and saturation headroom instead of on this number.\n"),
        best_n, runner_n))
    }
  }

  # Median is robust to exactly the far-end blow-up that says which count
  # degrades gracefully - so say when the two disagree.
  by_worst <- eligible %>% arrange(worst_resolution_mm)
  if (as.character(by_worst$n_magnets[1]) != best_n) {
    cat(sprintf(
      paste0("  NOTE: ranking on worst-case resolution instead would pick %s ",
             "magnet(s) (%.3f mm worst, vs %.3f mm for %s). Prefer that count ",
             "if consistency across the whole range matters more than the ",
             "typical case.\n"),
      as.character(by_worst$n_magnets[1]), by_worst$worst_resolution_mm[1],
      eligible$worst_resolution_mm[1], best_n))
  }
}
```

# EXPORT

## Coefficients

The file written here is what `gaping_analysis.Rmd` reads to convert
logged field magnitude into gape distance. `uid` travels with the
coefficients so the pairing can never be applied to the wrong board's
data.

The **background vector** travels with them too, as `bg_x_uT`, `bg_y_uT`
and `bg_z_uT`. That matters: these curves were fit to
background-corrected field, so anything applying them has to correct its
own readings the same way first, or it is feeding the curve a different
quantity than the one it was built from. Exporting the vector is what
makes that possible downstream. (With `subtract_background = FALSE` the
three columns are written as zeros, which makes the correction
downstream a no-op, as it should be.)

```{r export-coefficients}
coef_out <- coefficients %>%
  left_join(model_stats %>% select(n_magnets, model, n_points, rmse_mm, max_err_mm, aic),
            by = c("n_magnets", "model")) %>%
  left_join(best_models %>% select(n_magnets, best_model), by = "n_magnets") %>%
  left_join(
    step_summary %>%
      group_by(n_magnets) %>%
      summarise(fit_min_distance_mm = min(distance_mm),
                fit_max_distance_mm = max(distance_mm),
                fit_min_field_uT    = min(mean_uT),
                fit_max_field_uT    = max(mean_uT), .groups = "drop"),
    by = "n_magnets"
  ) %>%
  mutate(
    unit          = unit_label,
    rp2040_uid    = expected_uid,
    is_best_model = model == best_model,
    calibrated_on = Sys.Date(),
    .before       = everything()
  ) %>%
  # The ambient vector these curves were fit against. Without it downstream,
  # the curve gets applied to uncorrected field and quietly reads the wrong
  # distance - see the note above.
  mutate(
    bg_x_uT = background$bx,
    bg_y_uT = background$by,
    bg_z_uT = background$bz
  ) %>%
  select(-best_model)

coef_path <- file.path(output_dir,
                       sprintf("calibration-coefficients_%s.csv", unit_label))
write_csv(coef_out, coef_path)

cat(sprintf("Wrote %s\n", coef_path))
coef_out %>%
  mutate(across(where(is.numeric), ~ signif(.x, 5))) %>%
  kable(caption = "Exported calibration coefficients")
```

## Per-hold summary

```{r export-summary}
summary_path <- file.path(output_dir,
                          sprintf("calibration-hold-summary_%s.csv", unit_label))

step_summary %>%
  mutate(unit = unit_label, rp2040_uid = expected_uid,
         .before = everything()) %>%
  write_csv(summary_path)

cat(sprintf("Wrote %s\n", summary_path))
```

## Using the curve

```{r predict-demo}
predict_distance <- function(field_uT, a, b, c = 0) {
  a * field_uT^b + coalesce(c, 0)
}

demo_magnets <- recommended_n_magnets
demo <- coef_out %>% filter(n_magnets == demo_magnets, is_best_model)

tibble(field_uT = signif(exp(seq(log(demo$fit_min_field_uT),
                                 log(demo$fit_max_field_uT), length.out = 8)), 4)) %>%
  mutate(distance_mm = round(predict_distance(field_uT, demo$a, demo$b, demo$c), 2)) %>%
  kable(caption = sprintf("%s, %s magnet(s), %s model: field -> distance",
                          unit_label, demo_magnets, demo$model))
```

Two rules for using the exported CSV:

1.  **Select the row by `n_magnets`, not just by unit.** The
    coefficients are keyed on `(rp2040_uid, n_magnets, model)` precisely
    because the magnet count is part of the calibration. Whichever
    magnet count an oyster was deployed with is deployment metadata the
    logger does not record - it has to be written down at deployment and
    carried alongside the `gapelog_NNN.csv`, or there is no way to know
    which row applies.
2.  **Stay inside the fitted range** (`fit_min_field_uT` to
    `fit_max_field_uT`). A power law extrapolates confidently and
    wrongly outside the range it was fit on.
3.  **Subtract the background first.** These coefficients map
    *background-corrected* field to distance. Before applying them,
    subtract `bg_x_uT` / `bg_y_uT` / `bg_z_uT` from the logged axes and
    recompute magnitude from the corrected axes - per axis, not off the
    magnitude. `gaping_analysis.Rmd` does this for you.

# NOTES

-   **Per unit, not per magnet count.** Magnet count is a grouping
    variable here, not a separate notebook, so the three counts are
    fitted and compared side by side in one place - which is the
    comparison that decides how the units get deployed. Separate *files*
    per magnet count were considered and rejected; separate *rows* in
    the exported coefficients are what actually matters, and those are
    keyed on `n_magnets`.
-   **Magnet count must travel with the deployment.** Different oysters
    may need different magnet counts (shell thickness at the leading
    edge, animal size), and the logger records no magnet count - only
    the board UID. Note it down per deployment; without it the right
    calibration row cannot be chosen.
-   **The UID identifies the board, not the sensor.** If a TLV493D is
    moved to a different Adalogger, these coefficients follow the
    *sensor*, and the pairing has to be re-established by hand.
-   **Re-calibrate after any physical change** to the magnet, its
    mounting, or the sensor-magnet geometry. The curve is specific to
    that combination.

# REFERENCES
