---
title: "Oyster Gaping - Magnetometer Data Analysis"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
---

# OVERVIEW

This notebook reads the CSV log(s) produced by the CircuitPython datalogger
running on the Feather RP2040 Adalogger (`gapelog_001.csv`,
`gapelog_002.csv`, ... on the SD card), reconstructs real-world timestamps,
detects gape events, and plots the resulting time series.

**Data format on disk (wide form):** one row per sample, with the three
simultaneous magnetic field components as columns:

| column      | meaning                                                  |
|-------------|----------------------------------------------------------|
| `elapsed_s` | seconds since that log file's session started            |
| `x_uT`      | magnetic field, X axis, microtesla                       |
| `y_uT`      | magnetic field, Y axis, microtesla                       |
| `z_uT`      | magnetic field, Z axis, microtesla                       |

Wide form is kept on the device and on disk because X/Y/Z are simultaneous
components of one field vector, not independent long-form observations -
computing field magnitude, per-axis deltas, etc. all need all three values
from the same row. We pivot to long form only for specific plots below.

**One file per power-on.** The logger starts a new numbered file every time
the board boots, so each `gapelog_NNN.csv` is exactly one session. Session
identity comes from the filename - no need to infer reboots from the data.

**No onboard real-time clock.** The board has no battery-backed RTC, so it
records `elapsed_s` (seconds since that session started) rather than a
wall-clock date/time. Supply one real-world start time per session in the
SETUP section below to reconstruct absolute timestamps.

# SETUP

## Libraries

```{r libraries, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 5
)

library(tidyverse)
library(lubridate)
library(knitr)
```

## Analysis parameters

All user-configurable values are assigned here - edit this chunk for your
deployment and nothing else below should need changing.

```{r parameters}
# ---- Data location -------------------------------------------------------
# Directory holding the CSV files copied off the SD card, and the filename
# pattern to match. Every matching file is treated as one logging session.
data_dir      <- "."
file_pattern  <- "^gapelog_[0-9]+\\.csv$"

# ---- Deployment start times ---------------------------------------------
# One real-world start date/time per log file, in filename order
# (gapelog_001.csv, gapelog_002.csv, ...). Record these in the field at the
# moment you apply power to the board. Format "YYYY-MM-DD HH:MM:SS".
#
# If you leave this as NULL, timestamps stay RELATIVE (elapsed seconds) and
# every plot/metric below is still computed - just on relative time.
session_start_times <- NULL
# session_start_times <- c(
#   "2026-09-01 08:00:00",
#   "2026-09-03 14:22:00"
# )

deployment_tz <- "America/Los_Angeles"

# ---- Gape event detection -----------------------------------------------
# The magnet sits on one valve, the magnetometer on the other: as the shell
# OPENS the magnet moves away and field magnitude FALLS. So a gape event is
# a sustained DROP in magnitude below a threshold.
#
# The threshold can be set three ways:
#   "noise"    - computed per session from the spread (SD) of the
#                closed-shell baseline itself: threshold = closed_mean -
#                n_sd_threshold * closed_sd. This ties the cutoff to how
#                noisy that particular session's signal actually is, so a
#                "gape" is a drop that's statistically distinguishable from
#                background noise, not an arbitrary fractional drop.
#                RECOMMENDED DEFAULT.
#   "relative" - computed per session as a fixed fractional drop below the
#                closed-shell baseline value (kept for comparison /
#                backward compatibility - doesn't account for how noisy the
#                signal is).
#   "absolute" - a fixed microtesla value you supply.
threshold_method <- "noise"

# Used for both "noise" and "relative" methods: the quantile of magnitude
# taken as representative of "fully closed" (magnet closest to sensor).
baseline_quantile <- 0.90

# Used when threshold_method == "noise":
# The closed-baseline mean/SD are estimated from samples at/above
# `baseline_quantile`. n_sd_threshold sets how many SDs below that mean a
# sample must fall to count as "open" - larger values are more conservative
# (fewer false-positive events from sensor noise), smaller values are more
# sensitive. 3 SD is a common starting point for "clearly not noise".
n_sd_threshold <- 3

# Used when threshold_method == "relative":
gape_drop_frac <- 0.10   # open if magnitude < baseline * (1 - this)

# Used when threshold_method == "absolute":
gape_threshold_uT <- 40

# ---- Event filtering ----------------------------------------------------
# Reject blips and bridge momentary re-closures, so sensor noise doesn't
# inflate the event count.
min_event_duration_s <- 30   # discard "open" runs shorter than this
min_gap_between_s    <- 30   # merge two events separated by less than this
```

# DATA IMPORT

## Locate log files

```{r locate-files}
log_files <- list.files(
  path = data_dir, pattern = file_pattern, full.names = TRUE
) %>%
  sort()

stopifnot("No log files matched `file_pattern` in `data_dir`." =
            length(log_files) > 0)

tibble(
  file     = basename(log_files),
  size_kb  = round(file.size(log_files) / 1024, 1)
) %>%
  kable(caption = "Log files found")
```

## Read and combine

```{r read-data}
raw <- log_files %>%
  set_names(basename(.)) %>%
  map(
    read_csv,
    col_types = cols(
      elapsed_s = col_double(),
      x_uT      = col_double(),
      y_uT      = col_double(),
      z_uT      = col_double()
    )
  ) %>%
  list_rbind(names_to = "source_file") %>%
  mutate(session_id = dense_rank(source_file), .after = source_file)

glimpse(raw)
```

## Reconstruct timestamps

Real time = that session's start time + `elapsed_s`. If
`session_start_times` is `NULL`, `timestamp` stays `NA` and all downstream
timing uses `elapsed_s` instead.

```{r reconstruct-timestamps}
n_sessions <- n_distinct(raw$session_id)

use_absolute_time <- !is.null(session_start_times) &&
  length(session_start_times) == n_sessions

if (!is.null(session_start_times) && !use_absolute_time) {
  warning(sprintf(
    paste("Found %d log file(s) but %d entry/entries in",
          "`session_start_times`. Falling back to RELATIVE time for every",
          "session - supply exactly one start time per log file in the",
          "SETUP chunk to get absolute timestamps."),
    n_sessions, length(session_start_times)
  ))
}

if (use_absolute_time) {
  session_starts <- tibble(
    session_id    = seq_len(n_sessions),
    session_start = ymd_hms(session_start_times, tz = deployment_tz)
  )
} else {
  session_starts <- tibble(
    session_id    = seq_len(n_sessions),
    session_start = as.POSIXct(NA, tz = deployment_tz)
  )
}

gaping <- raw %>%
  left_join(session_starts, by = "session_id") %>%
  mutate(
    timestamp = session_start + elapsed_s,
    date      = as_date(timestamp),
    time      = format(timestamp, "%H:%M:%S")
  )

# Require EVERY session to have a real start time before plotting on a
# datetime axis - a partial mix would render some panels on real time and
# others on nothing, which is easy to misread.
has_real_time <- !any(is.na(gaping$timestamp))

cat(sprintf(
  "%d session(s), %d total samples. Absolute timestamps: %s\n",
  n_sessions, nrow(gaping),
  if (has_real_time) "YES" else "NO (relative elapsed_s only)"
))
```

# DERIVED METRICS

## Field magnitude

Field magnitude `sqrt(x^2 + y^2 + z^2)` is the primary gape proxy: it is
rotation-invariant, so it doesn't depend on exactly how the sensor is
oriented relative to the magnet.

```{r field-magnitude}
gaping <- gaping %>%
  mutate(field_magnitude_uT = sqrt(x_uT^2 + y_uT^2 + z_uT^2))

gaping %>%
  select(x_uT, y_uT, z_uT, field_magnitude_uT) %>%
  summary()
```

## Sampling interval

```{r sampling-interval}
gaping <- gaping %>%
  group_by(session_id) %>%
  arrange(elapsed_s, .by_group = TRUE) %>%
  mutate(sample_dt_s = elapsed_s - lag(elapsed_s)) %>%
  ungroup()

gaping %>%
  group_by(session_id, source_file) %>%
  summarise(
    n_samples     = n(),
    duration_min  = (max(elapsed_s) - min(elapsed_s)) / 60,
    median_dt_s   = median(sample_dt_s, na.rm = TRUE),
    max_gap_s     = max(sample_dt_s, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  kable(digits = 2, caption = "Per-session sampling summary")
```

# GAP EVENT DETECTION

## Threshold and open/closed state

The `"noise"` method characterizes the closed-shell baseline as a
distribution, not a single value: samples at/above `baseline_quantile` are
taken as "closed", and their mean and SD describe that session's baseline
noise. The threshold then sits `n_sd_threshold` SDs below that mean, so a
gape event is a drop that's statistically distinct from background noise -
not just a fixed percentage below baseline, which says nothing about
whether that drop is bigger than the sensor's own jitter.

```{r classify-state}
gaping <- gaping %>%
  group_by(session_id) %>%
  mutate(
    baseline_uT    = quantile(field_magnitude_uT, baseline_quantile, na.rm = TRUE),
    closed_mean_uT = mean(field_magnitude_uT[field_magnitude_uT >= baseline_uT], na.rm = TRUE),
    closed_sd_uT   = sd(field_magnitude_uT[field_magnitude_uT >= baseline_uT], na.rm = TRUE),
    threshold_uT = if (threshold_method == "noise") {
      closed_mean_uT - n_sd_threshold * closed_sd_uT
    } else if (threshold_method == "relative") {
      baseline_uT * (1 - gape_drop_frac)
    } else if (threshold_method == "absolute") {
      gape_threshold_uT
    } else {
      NA_real_
    },
    is_open = field_magnitude_uT < threshold_uT
  ) %>%
  ungroup()

stopifnot("`threshold_method` must be one of \"noise\", \"relative\", \"absolute\"." =
            !any(is.na(gaping$threshold_uT)))

gaping %>%
  group_by(session_id, source_file) %>%
  summarise(
    closed_mean_uT   = first(closed_mean_uT),
    closed_sd_uT     = first(closed_sd_uT),
    threshold_uT     = first(threshold_uT),
    pct_samples_open = 100 * mean(is_open),
    .groups = "drop"
  ) %>%
  kable(digits = 2, caption = "Detection threshold per session")
```

## Build the event table

Contiguous runs of `is_open` become candidate events; runs shorter than
`min_event_duration_s` are dropped, and events separated by less than
`min_gap_between_s` are merged into one.

```{r build-events}
# 1. Contiguous runs of the same open/closed state, per session.
runs <- gaping %>%
  group_by(session_id) %>%
  mutate(run_id = cumsum(is_open != lag(is_open, default = first(is_open)))) %>%
  group_by(session_id, source_file, run_id, is_open) %>%
  summarise(
    start_s   = min(elapsed_s),
    end_s     = max(elapsed_s),
    n_samples = n(),
    mean_magnitude_uT = mean(field_magnitude_uT),
    min_magnitude_uT  = min(field_magnitude_uT),
    .groups   = "drop"
  ) %>%
  mutate(duration_s = end_s - start_s)

# 2. Keep open runs long enough to count as real events.
events <- runs %>%
  filter(is_open, duration_s >= min_event_duration_s) %>%
  arrange(session_id, start_s)

# 3. Merge events separated by a shorter closure than min_gap_between_s.
events <- events %>%
  group_by(session_id) %>%
  mutate(
    gap_before_s = start_s - lag(end_s),
    new_event    = is.na(gap_before_s) | gap_before_s >= min_gap_between_s,
    event_group  = cumsum(new_event)
  ) %>%
  group_by(session_id, source_file, event_group) %>%
  summarise(
    start_s           = min(start_s),
    end_s             = max(end_s),
    # Weighted mean must be computed BEFORE n_samples is collapsed to a
    # single sum, otherwise the weights vector no longer matches the values.
    mean_magnitude_uT = weighted.mean(mean_magnitude_uT, n_samples),
    min_magnitude_uT  = min(min_magnitude_uT),
    n_samples         = sum(n_samples),
    .groups = "drop"
  ) %>%
  group_by(session_id) %>%
  mutate(
    event_id    = row_number(),
    duration_s  = end_s - start_s,
    duration_min = duration_s / 60
  ) %>%
  ungroup() %>%
  left_join(session_starts, by = "session_id") %>%
  mutate(
    start_time = session_start + start_s,
    end_time   = session_start + end_s
  )

cat(sprintf("Detected %d gape event(s) across %d session(s).\n",
            nrow(events), n_sessions))

events %>%
  select(session_id, event_id, start_s, end_s, duration_s, duration_min,
         mean_magnitude_uT, min_magnitude_uT, n_samples) %>%
  kable(digits = 2, caption = "Detected gape events")
```

## Event frequency

```{r event-frequency}
event_frequency <- events %>%
  group_by(session_id, source_file) %>%
  summarise(n_events = n(), .groups = "drop") %>%
  left_join(
    gaping %>%
      group_by(session_id) %>%
      summarise(
        recording_hr = (max(elapsed_s) - min(elapsed_s)) / 3600,
        .groups = "drop"
      ),
    by = "session_id"
  ) %>%
  mutate(
    events_per_hr  = n_events / recording_hr,
    events_per_day = events_per_hr * 24
  )

event_frequency %>%
  kable(digits = 3, caption = "Gape event frequency per session")
```

## Event durations

```{r event-durations}
duration_stats <- events %>%
  group_by(session_id, source_file) %>%
  summarise(
    n_events        = n(),
    mean_duration_s = mean(duration_s),
    sd_duration_s   = sd(duration_s),
    min_duration_s  = min(duration_s),
    median_duration_s = median(duration_s),
    max_duration_s  = max(duration_s),
    total_open_s    = sum(duration_s),
    .groups = "drop"
  ) %>%
  left_join(
    gaping %>%
      group_by(session_id) %>%
      summarise(
        recording_s = max(elapsed_s) - min(elapsed_s), .groups = "drop"
      ),
    by = "session_id"
  ) %>%
  mutate(pct_time_open = 100 * total_open_s / recording_s)

duration_stats %>%
  kable(digits = 2, caption = "Gape event duration statistics (seconds)")
```

## Time between events

Inter-event interval = time from the **end** of one event to the **start**
of the next (i.e. how long the animal stayed closed between gapes).

```{r inter-event-intervals}
intervals <- events %>%
  group_by(session_id, source_file) %>%
  arrange(start_s, .by_group = TRUE) %>%
  mutate(interval_s = start_s - lag(end_s)) %>%
  filter(!is.na(interval_s)) %>%
  ungroup()

interval_stats <- intervals %>%
  group_by(session_id, source_file) %>%
  summarise(
    n_intervals      = n(),
    mean_interval_s  = mean(interval_s),
    sd_interval_s    = sd(interval_s),
    shortest_interval_s = min(interval_s),
    median_interval_s   = median(interval_s),
    longest_interval_s  = max(interval_s),
    .groups = "drop"
  )

if (nrow(interval_stats) > 0) {
  interval_stats %>%
    kable(digits = 2,
          caption = "Time between consecutive gape events (seconds)")
} else {
  cat("Fewer than 2 events per session - no inter-event intervals to report.\n")
}
```

## Combined summary

```{r combined-summary}
gape_summary <- event_frequency %>%
  select(session_id, source_file, n_events, recording_hr,
         events_per_hr, events_per_day) %>%
  left_join(
    duration_stats %>%
      select(session_id, mean_duration_s, min_duration_s, max_duration_s,
             total_open_s, pct_time_open),
    by = "session_id"
  ) %>%
  left_join(
    interval_stats %>%
      select(session_id, mean_interval_s, shortest_interval_s,
             longest_interval_s),
    by = "session_id"
  )

gape_summary %>%
  kable(digits = 2, caption = "Gape event summary, all metrics by session")
```

# PLOTS

## Field magnitude with detected events

Shaded bands are detected gape events; the dashed line is the detection
threshold.

```{r plot-magnitude-events}
x_var  <- if (has_real_time) "timestamp" else "elapsed_s"
x_lab  <- if (has_real_time) "Time" else "Elapsed time (s)"

p_magnitude <- ggplot(gaping, aes(x = .data[[x_var]], y = field_magnitude_uT))

if (nrow(events) > 0) {
  event_bands <- events %>%
    mutate(
      xmin = if (has_real_time) start_time else start_s,
      xmax = if (has_real_time) end_time   else end_s
    )
  p_magnitude <- p_magnitude +
    geom_rect(
      data = event_bands,
      aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
      inherit.aes = FALSE, fill = "steelblue", alpha = 0.20
    )
}

p_magnitude <- p_magnitude +
  geom_line(linewidth = 0.3) +
  geom_hline(
    aes(yintercept = threshold_uT), linetype = "dashed", color = "firebrick"
  ) +
  facet_wrap(~source_file, scales = "free_x", ncol = 1) +
  labs(
    title = "Field magnitude over time, with detected gape events shaded",
    subtitle = "Magnitude falls as the shell opens (magnet moves away from sensor)",
    x = x_lab, y = "Field magnitude (uT)"
  ) +
  theme_minimal()

p_magnitude
```

## Per-axis time series

Pivoting to long form here is purely a plotting convenience - it is not how
the data is stored.

```{r plot-per-axis}
gaping_long <- gaping %>%
  select(session_id, source_file, all_of(x_var), x_uT, y_uT, z_uT) %>%
  pivot_longer(
    cols = c(x_uT, y_uT, z_uT), names_to = "axis", values_to = "field_uT"
  )

p_axes <- ggplot(gaping_long, aes(x = .data[[x_var]], y = field_uT, color = axis)) +
  geom_line(linewidth = 0.3) +
  facet_grid(axis ~ source_file, scales = "free") +
  labs(
    title = "Magnetic field components over time",
    x = x_lab, y = "Field (uT)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

p_axes
```

## Event duration distribution

```{r plot-duration-distribution}
if (nrow(events) > 0) {
  p_durations <- ggplot(events, aes(x = duration_min)) +
    geom_histogram(bins = 30, fill = "steelblue", color = "white") +
    facet_wrap(~source_file, ncol = 1) +
    labs(
      title = "Distribution of gape event durations",
      x = "Event duration (minutes)", y = "Count of events"
    ) +
    theme_minimal()
  p_durations
} else {
  cat("No events detected - nothing to plot.\n")
}
```

## Event duration over time

Shows whether gape bouts get longer or shorter across the deployment.

```{r plot-duration-over-time}
if (nrow(events) > 0) {
  p_dur_time <- events %>%
    mutate(x_pos = if (has_real_time) start_time else start_s) %>%
    ggplot(aes(x = x_pos, y = duration_min)) +
    geom_point(size = 2, color = "steelblue") +
    geom_line(linewidth = 0.3, alpha = 0.5, color = "steelblue") +
    facet_wrap(~source_file, scales = "free_x", ncol = 1) +
    labs(
      title = "Gape event duration over the course of the deployment",
      x = x_lab, y = "Event duration (minutes)"
    ) +
    theme_minimal()
  p_dur_time
} else {
  cat("No events detected - nothing to plot.\n")
}
```

## Inter-event intervals

```{r plot-intervals}
if (nrow(intervals) > 0) {
  p_intervals <- ggplot(intervals, aes(x = interval_s / 60)) +
    geom_histogram(bins = 30, fill = "darkorange", color = "white") +
    facet_wrap(~source_file, ncol = 1) +
    labs(
      title = "Time between consecutive gape events",
      x = "Interval between events (minutes)", y = "Count"
    ) +
    theme_minimal()
  p_intervals
} else {
  cat("Fewer than 2 events per session - no intervals to plot.\n")
}
```

## Events per hour by session

```{r plot-event-rate}
p_rate <- ggplot(event_frequency, aes(x = source_file, y = events_per_hr)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Gape event frequency by session",
    x = "Log file (session)", y = "Events per hour"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_rate
```

# NOTES / NEXT STEPS

- **Threshold calibration.** The `noise` and `relative` methods both assume
  the animal is closed for most of the recording (the top
  `1 - baseline_quantile` fraction of samples, default 10%, is treated as
  "closed" and used to characterize the baseline). If an animal gapes more
  than that, the baseline estimate - and for `noise`, the SD used to set
  the threshold - gets contaminated by open samples and the threshold
  drifts too low. If that assumption breaks, characterize known
  open/closed magnitudes on the bench and switch to
  `threshold_method = "absolute"`.
- **Noise-based threshold (`n_sd_threshold`).** The default of 3 SD below
  the closed-baseline mean is a starting point, not a calibrated value -
  check the `closed_sd_uT` column in the per-session table. A very small SD
  (a clean, low-noise sensor) with 3 SD may flag drops that are real but
  small; a very large SD (noisy signal, or a baseline window that already
  contains some open samples) may require a larger `n_sd_threshold` to
  avoid false positives. Vary it and confirm the detected event count and
  `pct_samples_open` are stable rather than an artifact of one setting.
- **Event filtering.** `min_event_duration_s` and `min_gap_between_s` set
  how aggressively blips are rejected and near-adjacent gapes are merged.
  Vary them and re-render to check the event count is stable rather than an
  artifact of one particular setting.
- **Timestamps.** Record the real power-on time for each `gapelog_NNN.csv`
  in the field and enter them in `session_start_times`; without them all
  timing is relative to the start of each file.
- **Adding an RTC** (PCF8523 / DS3231 breakout) would remove the manual
  start-time bookkeeping entirely, since the logger could then write real
  date/time to each row.
