---
title: "20211006_cbai_SNP_stats"
author: "Sam White"
date: "10/5/2021"
output: html_document
---

```{r include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Explore and extract SNP data from *C.bairdi* transcritpome assembly v3.1

REQUIRES Linux-based system to run all cells properly; some cells will not work on Mac OS!

REQUIRES the following programs:

- [bcftools](https://github.com/samtools/bcftools)

- [samtools](https://github.com/samtools/samtools)

- [seqtk](https://github.com/lh3/seqtk)

REQUIRES the following R libraries and dependencies:

- `goslim_generic.obo` Downloaded from http://geneontology.org/docs/go-subset-guide/
  - then i moved it to the R library for GSEABase in the extdata folder in addition to using the command here - I think they're both required.

- `GSEABase` (BioConductor)

- `tidyverse`


### Display system info

```{bash}
echo "TODAY'S DATE:"
date
echo "------------"
echo ""
#Display operating system info
lsb_release -a
echo ""
echo "------------"
echo "HOSTNAME: "; hostname 
echo ""
echo "------------"
echo "Computer Specs:"
echo ""
lscpu
echo ""
echo "------------"
echo ""
echo "Memory Specs"
echo ""
free -mh
```


### User-defined bash variables

Set program paths for your own computer.

Variables are saved to a "dot file" and that file needs to be sourced in each Bash chunk to have access to the Bash variables
across Bash chunks.

```{bash echo = TRUE}
{
echo "# CPU threads"
echo 'export threads=8'
echo ""
echo "# Programs"
echo 'export seqtk="/home/sam/programs/seqtk-1.3/seqtk"'
echo 'export bcftools="/home/sam/programs/bcftools-1.13/bcftools"'
echo 'export samtools="/home/sam/programs/samtools-1.12/samtools"'
echo ""
} > .rvars
```


### Input/output files variables
```{bash echo = TRUE}
{
echo "# SNP coverage"
echo 'export SNP_coverage=10'
echo ""
echo "# SNP quality"
echo 'export SNP_quality=30'
echo ""

echo "# Input files"
echo ""
echo "## Transcriptome assembly"
echo 'export orig_fasta_url_dir="https://owl.fish.washington.edu/halfshell/genomic-databank"'
echo 'export transcriptome_fasta="cbai_transcriptome_v3.1.fasta"'
echo ""
echo "## Transcriptome md5 checksum"
echo 'export transcriptome_fasta_md5="aeec8ffbf8fa44fb1750caee6abaf68a"'
echo ""
echo "## Transcriptome GO terms"
echo 'export cbai_v3_1_GO_url="https://gannet.fish.washington.edu/Atumefaciens/20200828_cbai_trinotate_transcriptome-v3.1"'
echo 'export cbai_v3_1_GO="20200828.cbai_transcriptome_v3.1.fasta.trinotate.go_annotations.txt"'
echo ""
echo "## VCF with variant calls"
echo 'export orig_vcf_url_dir="https://gannet.fish.washington.edu/Atumefaciens/20210909-cbai-bcftools-snp_calling"'
echo 'export orig_vcf="cbai_v3.1-SNPS.vcf"'
echo ""

echo "# Output files"
echo 'export transcriptome_SNPS_fasta="cbai_transcriptome_v3.1_SNPs-${SNP_quality}Q-${SNP_coverage}x.fasta"'
echo 'export contigs_list="cbai_v3.1-SNPS_${SNP_quality}Q-${SNP_coverage}x_contig-IDs.txt"'
echo 'export genes_list="cbai_v3.1-SNPS_${SNP_quality}Q-${SNP_coverage}x_gene-IDs.txt"'
echo 'export vcf_filtered="cbai_v3.1-SNPS-${SNP_quality}Q-${SNP_coverage}x.vcf"'
echo 'export genes_GO_list"=cbai_v3.1-SNPS_${SNP_quality}Q-${SNP_coverage}x_GO.tab"'
echo 'export flattened_GO="cbai_v3_1-SNPS_${SNP_quality}Q-${SNP_coverage}x_GO.flattened-go.txt"'
echo ""

echo "# Print formatting"
echo 'export line="-------------------------------------------------------------------------------------------------"'
echo ""
} >> .rvars

```


### Confirm variables are accessible.
```{bash echo = TRUE}
# Confirm contents of .rvars
cat .rvars

# Load contents of .rvars into the environment
source .rvars

echo ""

echo "Confirm variables are accessible."
echo "Checking the variable \$line:"
echo "${line}"
```

### Get VCF
```{bash echo = TRUE}
# Load contents of .rvars into the environment
source .rvars

# Download with wget. Use --no-check-certificate to avoid issues with Gannet certificate
# Use --quiet option to prevent wget output from printing too many lines to notebook
wget --continue --no-check-certificate --quiet ${orig_vcf_url_dir}/${orig_vcf} \
--directory-prefix ../data

wget --continue --no-check-certificate --quiet ${orig_vcf_url_dir}/checksums.md5 \
--directory-prefix ../data

# Confirm checksum for transcriptome FastA is good
cd ../data
md5sum --check ../data/checksums.md5 | grep "${orig_vcf}"
```

### Get transcriptome
```{bash echo = TRUE}
# Load contents of .rvars into the environment
source .rvars

# Download with wget. Use --no-check-certificate to avoid issues with Gannet certificate
# Use --quiet option to prevent wget output from printing too many lines to notebook
wget --continue --no-check-certificate --quiet ${orig_fasta_url_dir}/${transcriptome_fasta} \
--directory-prefix ../data


# Confirm checksum for transcriptome FastA is good
# Uses grep to highlight the desired file.
if [ "$(md5sum ../data/${transcriptome_fasta} | awk '{print $1}')" = "${transcriptome_fasta_md5}" ]; then echo "Checksums match"; fi
```

### Get transcriptome GO annotations file
```{bash echo = TRUE}
# Load contents of .rvars into the environment
source .rvars

# Download with wget. Use --no-check-certificate to avoid issues with Gannet certificate
# Use --quiet option to prevent wget output from printing too many lines to notebook
wget --continue --no-check-certificate --quiet ${cbai_v3_1_GO_url}/${cbai_v3_1_GO} \
--directory-prefix ../data

echo ""

head ../data/${cbai_v3_1_GO}
```

### Inspect original VCF
```{bash echo = TRUE}
# Load contents of .rvars into the environment
source .rvars

echo "'head' view of ${orig_vcf}:"
echo ""
head "../data/${orig_vcf}"
echo ""
echo "End of 'head' view of ${orig_vcf}"
echo ""
echo "${line}"
echo ""

echo "VCF header info:"
echo ""

# Capture first line of header (skipping list of contigs)
begin=$("${bcftools}" view --header-only ../data/${orig_vcf} \
| grep --line-number "##ALT=<ID" \
| awk -F":" '{print $1}')

# Caputure last line of header
end=$("${bcftools}" view --header-only ../data/${orig_vcf} | wc -l)

# Use sed to print range of lines
sed --quiet "${begin},${end} p" ../data/${orig_vcf}
echo ""
echo "${line}"
echo ""

echo "List of samples in VCF:"
echo ""
${bcftools} query --list-samples ../data/${orig_vcf}
```

### Subset VCF to minimum coverage and quality
```{bash echo = TRUE}
# Load contents of .rvars into the environment
source .rvars

# Subset VCF to only SNPs with ${SNP_coverage}x raw read coverage
# and quality >= ${SNP_qual}
"${bcftools}" filter \
--include "TYPE='snp' & MIN(DP)>=${SNP_coverage} & QUAL>=${SNP_quality}" \
--threads ${threads} \
../data/${orig_vcf} \
> ../analyses/${vcf_filtered}
```

### Inspect filtered VCF
```{bash echo = TRUE}
# Load contents of .rvars into the environment
source .rvars

echo "'head' view of ${vcf_filtered}:"
echo ""
head "../analyses/${vcf_filtered}"
echo ""
echo "End of 'head' view of ${vcf_filtered}"
echo ""
echo "${line}"
echo ""

echo "VCF header info:"
echo ""

# Capture first line of header (skipping list of contigs)
begin=$("${bcftools}" view --header-only ../analyses/${vcf_filtered} \
| grep --line-number "##ALT=<ID" \
| awk -F":" '{print $1}')

# Caputure last line of header
end=$("${bcftools}" view --header-only ../analyses/${vcf_filtered} | wc -l)

# Use sed to print range of lines
 sed --quiet "${begin},${end} p" ../analyses/${vcf_filtered}

echo ""
echo "${line}"
echo ""

echo "List of samples in VCF:"
echo ""
${bcftools} query --list-samples ../analyses/${vcf_filtered}
```


### Extract Transcripts Having SNPs with `${SNP_coverage}x` Read Coverage and Quality >= `${SNP_qual}`

The resulting FastA is useful to use in IGV, so that we don't have to deal with browsing contigs with no variants.
```{bash echo = TRUE}
# Load contents of .rvars into the environment
source .rvars

# List of FastA IDs with ${SNP_coverage}x SNP coverage
# Uses awk to skip all lines beginning with '#'
awk '/^[^#]/{print $1}' ../analyses/${vcf_filtered} \
| sort -u \
> ../analyses/${contigs_list}

# Use seqtk to generate a FastA from original transcriptome assembly and list of contigs with ${SNP_coverage}x SNP coverage
"${seqtk}" subseq ../data/${transcriptome_fasta} ../analyses/${contigs_list} > ../data/${transcriptome_SNPS_fasta}

# Generate FastA index file for new FastA
"${samtools}" faidx ../data/${transcriptome_SNPS_fasta}

ls -ltrh ../analyses

```



### Compare number of original transcripts with number of those with SNPs
```{bash echo = TRUE}
# Load contents of .rvars into the environment
source .rvars

for fasta in ../data/*.fasta
do
  grep --count --with-filename "^>" ${fasta}
done | column -t -s ":"
```


## SNP Stats

### Summary stats
```{bash echo = TRUE}
# Load contents of .rvars into the environment
source .rvars

# Shows summary stats for all samples
${bcftools} stats \
--samples - \
../analyses/${vcf_filtered} \
| head -n 31
```


#### Transitions/Transversions and Substitution types
```{bash echo = TRUE}
# Load contents of .rvars into the environment
source .rvars

${bcftools} stats \
--samples - \
../analyses/${vcf_filtered} \
| grep "ST"
```


### Individual sample stats
```{bash echo = TRUE}
# Load contents of .rvars into the environment
source .rvars

# Shows samples stats.
# Uses sed/column/tail to format output nicely
${bcftools} stats \
--samples - \
../analyses/${vcf_filtered} \
| grep "PSC" \
| sed 's/^# *//' \
| sed 's/average depth/avg_dp/' \
| column -t \
| tail -n 5
```



### Percentage of transcripts with SNPS
```{bash echo = TRUE}
# Load contents of .rvars into the environment
source .rvars

# Count number of transcripts in transcriptome assembly
transcripts_count=$(grep -c "^>" ../data/${transcriptome_fasta})
printf "%s\t%s\n" "Original transcriptome transcripts:" "${transcripts_count}"

# Count number of transcripts with SNPs
# Parses Trinity ID and counts unique Trinity IDs
snp_transcripts_count=$(awk '/^[^#]/{print $1}' ../analyses/${vcf_filtered} | sort -u| wc -l)
printf "%s\t%s\n" "Transcripts with SNPs:" "${snp_transcripts_count}"

echo ""
# Calculate percentage
printf "%s\t%s\n" "Percentage of transcripts with SNPs:" "$(bc <<< "scale=4; ( ${snp_transcripts_count} / ${transcripts_count} * 100)")"
```


### Get max/min/mean number of SNPs per transcript
```{bash echo = TRUE}
# Load contents of .rvars into the environment
source .rvars

awk '/^[^#]/{print $1}' ../analyses/${vcf_filtered} \
| sort \
| uniq -c \
| sort -k1n,1 \
| awk '{sum+=$1;cnt++;max=$1;min=cnt==1?$1:min} END{print "min="min, "max="max, "mean="sum/cnt}'
```


### Extract FastA IDs
Strips Trinity isoform designations from end of ID to match gene IDs in GO term annotation file
```{bash echo = TRUE}
# Load contents of .rvars into the environment
source .rvars

awk '/^[^#]/{print $1}' ../analyses/${vcf_filtered} \
| awk 'BEGIN { FS = "_"; OFS = "_" } {print $1, $2, $3, $4}' \
| sort -u \
> ../analyses/${genes_list}

wc -l ../analyses/"${genes_list}"

echo ""
echo "${line}"
echo ""

head ../analyses/"${genes_list}"
```


### Extract genes with SNPs from transcriptome annotation file
```{bash echo = TRUE}
# Load contents of .rvars into the environment
source .rvars

# Uses a list of Trinity gene IDs to extract gene IDs and corresponding GO accessions.
# Expects Trinotate annotations file as input.
# Uses sed to convert commas to tabs in preparation for subsequent "flattening"
while read -r line
do 
  grep "${line}" ../data/${cbai_v3_1_GO}
done < ../analyses/${genes_list} \
| sed $'s/,/\t/g' \
> ../analyses/${genes_GO_list}

# Check number of records
wc -l ../analyses/${genes_GO_list}

echo ""
echo "${line}"
echo ""

head ../analyses/${genes_GO_list}
```


### Flatten `${genes_GO_list}` to have one GO accession per line.
```{bash echo = TRUE}
# Load contents of .rvars into the environment
source .rvars

# Identify first field containing a GO term.
# Search file with grep for "GO:" and pipe to awk.
# Awk sets tab as field delimiter (-F'\t'), runs a for loop that looks for "GO:" (~/GO:/), and then prints the field number).
# Awk results are piped to sort, which sorts unique by number (-ug).
# Sort results are piped to head to retrieve the lowest value (i.e. the top of the list; "-n1").
begin_goterms=$(grep "GO:" ../analyses/"${genes_GO_list}" | awk -F'\t' '{for (i=1;i<=NF;i++) if($i ~/GO:/) print i}' | sort -ug | head -n1)

# Flatten GO terms annotation file.
# Expects tab-delimited input file where:
## First field (column) is Trinity gene IDs.
## Remaining fields (columns) are each individual GO accessions.
while read -r line
do
  # Capture maximum number of fields to handle differing number of GO terms.
  max_field=$(echo "$line" | awk -F'\t' '{print NF}')
  
  # Set which fields are "fixed" (i.e. Trinity gene IDs)
  fixed_fields=$(echo "$line" | cut -f1)
  
  # Identifies if a line has GO accessions,
  # reads them into an array,
  # and then prints the Trinity Id and single GO accession on each line.
  if (( "$max_field" < "$begin_goterms" )); then
    printf "%s\n" "$line"
  else
    # Set range of GO accessions for each line
    goterms=$(echo "$line" | cut -f"$begin_goterms"-"$max_field")
    # Set Internal Field Separator to a <tab> and read $goterms into array.
    IFS=$'\t' read -r -a array <<< "$goterms"
    # Loop through array and print tab-delimited file of Trinity ID and single GO accession.
    for element in "${!array[@]}"
      do printf "%s\t%s\n" "$fixed_fields" "${array[$element]}"
    done
  fi
done < ../analyses/"${genes_GO_list}" > ../analyses/"${flattened_GO}"

# Check number of records
wc -l ../analyses/${flattened_GO}

echo ""
echo "${line}"
echo ""

head ../analyses/${flattened_GO}
```

### Create file with path to `${flattened_GO}` file

Used to pass Bash variable to R.
```{bash echo = TRUE}
source .rvars
echo "../analyses/${flattened_GO}" > .flattened_GO
```


### Create R string with path to Bash variable, `${flattened_GO}`
```{r echo = TRUE}
string <- paste(readLines(".flattened_GO"), collapse=" ")
print(string)

```


```{r echo = TRUE}
library(GSEABase)
library(tidyverse)

# Script to retrieve GOslims from Trinity-based, EdgeR GOseq differential gene expression.
# Identifies enriched and depleted output files.
# Requires "goslim_generic.obo" from http://geneontology.org/docs/go-subset-guide/
  
  ## Get max number of fields
  # Needed to handle reading in file with different number of columns in each row
  max_fields <- max(na.omit((count.fields(string, sep = "\t", blank.lines.skip = TRUE))))
  
  ## Read in tab-delimited GOseq file
  # Use "max_fields" to populate all columns with a sequentially numbered header
  go_seqs <- read.table(string,
                        sep = "\t",
                        header = FALSE,
                        col.names = paste0("V",seq_len(max_fields)),
                        fill = TRUE)
  
  ## Grab just the individual GO terms from the "2nd" column)
  goterms <- as.character(go_seqs$V2)
  
  ### Use GSEA to map GO terms to GOslims
  
  ## Store goterms as GSEA object
  myCollection <- GOCollection(goterms)
  
  ## Use generic GOslim file to create a GOslim collection
  
  # I downloaded goslim_generic.obo from http://geneontology.org/docs/go-subset-guide/
  # then i moved it to the R library for GSEABase in the extdata folder
  # in addition to using the command here - I think they're both required.
  slim <- getOBOCollection("~/data/goslim_generic.obo")
  
  ## Map GO terms to GOslims and select Biological Processes group
  slims <- goSlim(myCollection, slim, "BP", verbose = TRUE)
  
  # Rename first column
  slims <- slims %>% rownames_to_column(var = "GOslim")
  
  ## Write output file
  write.csv(slims, file = "S10-SNPs_GO-GOslims.csv", quote = FALSE, row.names = FALSE)

# Remove GOslim accession for the generic "biological_process" term to improve visualization of other terms.
slims <- slims[slims$GOslim != "GO:0008150",]

# Create bar plot.
# "Open" PNG file for saving subsequent plot
pdf("../figures/S9-SNPs_GO-GOslims_barplot.pdf", height = 10, width = 12)

ggplot(data = slims, aes(x=slims$Percent, y=slims$Term)) +
  labs(title = "",
       caption = "Supplemental Figure 1. Barplot of percentages of gene ontology assignments to GOslims for transcripts containing at least on SNP.
       Excludes the generic \"biological_process\" GOslim (55.53% of all GO terms) to aid in visualization of other GOslim categories.",
       x = "Percent GO terms assigned to GOslim",
       y = "GOslim") +
  geom_bar(stat = "identity") +
  theme(plot.caption = element_text(hjust = 0)
  )

# Close PNG file
dev.off()
```